---
title: "Data Types & Objects"
output: html_document
---
  
## Vectors: The Building Block

Think of a vector as a list of information. It is the most simple type of object that R can have. One number, by itself, is a vector. You can also concatenate those single numbers together into a longer vector:

```{r}
first_vector <- c(1, 3, 5, 7)

first_vector
```

The vector we just created is a special type of vector: an *atomic* vector. Everything is the same type throughout the entire vector or it will be forced to be so. You can do cool things with vectors, like adding numbers to each *element* of the vector:

```{r}
first_vector + 10
```

All we have done is to use R like a fancy calculator! Take just a minute and try the following:

1. Take the `first_vector` object and perform some type of math operation. You can use symbols like: `*`, `/`, `+`,  `-`, `^`, `%/%`, `%%`

2. Whatever you did for step 1, create a new object called `second_vector`


### Types

#### Numeric

Numeric vectors are what would be considered *floating-point numbers* -- an overly-complicated way of saying that these numbers have decimals! Let's create a numeric vector:

```{r}
numeric_vector <- c(1.2, 2.2, 3.2, 4.2)

typeof(numeric_vector)
```

This returned `double`, which indicates that we are dealing with a "**double**-precision floating-point" number.

#### Integer

Just like you learned in elementary school, integers are whole numbers. 

Now we are going to see where things get a bit complicated.

What type would the following vector produce?

```{r}
test_vec <- c(10, 20, 30, 40, 50)
```

Now we can talk about conversion!

#### Strings

Strings are characters. A vector containing anything other than a number is a character:

```{r}
character_vector <- c("ABC", "abc", "1,000", "$100", 123)

character_vector

class(character_vector)
```


#### Factors

Ugh...a factor is a string with complication. Since R is a programming language based upon statistics, the notion of the factor is important. You can regard factors as categorical variables:

```{r}
factor_variable <- as.factor(c("freshman", "sophomore", "junior", "senior"))

factor_variable
```

This creates a weird problem. Some factors have no clear ordering to them -- factors like race, ethnicity, and gender -- but what we see above has a clear ordering to it. This is where the ordered factor comes in:

```{r}
ordered_factor <- ordered(factor_variable, 
                          levels = c("freshman", "sophomore", "junior", "senior"), 
                          labels = c("freshman", "sophomore", "junior", "senior"))
```



#### Logical

#### Specials

## Matrices

## Lists

## Data Frames

## Working With Data


