---
title: "Data Types & Objects"
output:
  html_document:
    theme: darkly
    highlight: breezedark
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "")
```
  
## Vectors: The Building Block

Think of a vector as a list of information. It is the most simple type of object that R can have. One number, by itself, is a vector (a vector of length 1). You can also concatenate (that is what the `c` function does) those single numbers together into a longer vector:

```{r}
first_vector <- c(1, 3, 5, 7)

first_vector
```

Note the assignment operator (you can also use an `=`, but people have strong opinions). On a Windows machine, you can insert the `<-` symbol with Alt + - and on a Mac it is option + -

The vector we just created is a special type of vector: an *atomic* vector. Everything is the same type throughout the entire vector or it will be forced to be so. You can do cool things with vectors, like adding numbers to each *element* of the vector:

```{r}
first_vector + 10
```

All we have done is to use R like a fancy calculator! Take just a minute and try the following:

1. Take the `first_vector` object and perform some type of math operation. You can use symbols like: `*`, `/`, `+`,  `-`, `^`, `%/%`, `%%`

2. Whatever you did for step 1, create a new object called `second_vector`

### Types

#### Numeric

Numeric vectors are what would be considered *floating-point numbers* -- an overly-complicated way of saying that these numbers have decimals! Let's create a numeric vector:

```{r}
numeric_vector <- c(1.2, 2.2, 3.2, 4.2)

typeof(numeric_vector)
```

This returned `double`, which indicates that we are dealing with a "**double**-precision floating-point" number.

#### Integer

Just like you learned in elementary school, integers are whole numbers. 

Now we are going to see where things get a bit complicated.

What type would the following vector produce?

```{r}
test_vec <- c(10, 20, 30, 40, 50)
```

Now we can talk about conversion!

#### Strings

Strings are characters. A vector containing anything other than a number is a character:

```{r}
character_vector <- c("ABC", "abc", "1,000", "$100", 123)

character_vector

class(character_vector)
```

It will take a while to warm up to strings, but the real magic comes when you dive into fun string operations!

#### Factors

Ugh...a factor is a string with complication. Since R is a programming language based upon statistics, the notion of the factor is important. You can regard factors as categorical variables:

```{r}
factor_variable <- as.factor(c("freshman", "sophomore", "junior", "senior"))

factor_variable
```

This creates a weird problem. Some factors have no clear ordering to them -- factors like race, ethnicity, and gender -- but what we see above has a clear ordering to it. This is where the ordered factor comes in:

```{r}
ordered_factor <- ordered(factor_variable, 
                          levels = c("freshman", "sophomore", "junior", "senior"), 
                          labels = c("freshman", "sophomore", "junior", "senior"))
```

#### Logical

Logical vectors contain TRUE and FALSE:

```{r}
logical_vector <- c(TRUE, FALSE, TRUE, FALSE)

logical_vector
```

Check this out!

```{r}
sum(logical_vector)
```

To use `T` and `F`?

#### Specials

You will encounter other values within your vectors. 

NA: This is a missing value

NaN: Not a number. R was expecting a number, but didn't get one.

Inf: Infinite. You'll see this happen with many different math operations (e.g., `cumsum`)

Null: Usually is an undefined value.

## Matrices

If we take a bunch of vectors, we can put them together into a special type of object called a `matrix`.

```{r}
matrix_example <- matrix(data = c(1:3, 4:6, 7:9), 
                         nrow = 3, 
                         ncol = 3)

matrix_example
```

One limitation you will see in a matrix is that every column must be of the same type.

Matrices are great for holding information and are one of the lighter ways of representing data. Another issue is that they are just a little trickier to work on.

## Lists

Now we are starting to get into some more interesting ideas. A list is another way to collect objects. However, we are now freed from the constraint that everything needs to be of the same type.

```{r}
list_example <- list(names = c("seth", "sharif"), 
                     ages = c(30, 30), 
                     height = c(76, 76))

list_example
```

It shouldn't come as a shock that a list can also contain lists:

```{r}
list_of_lists <- list(professors = list(names = c("seth", "sharif"), 
                                        ages = c(30, 30), 
                                        height = c(76, 76)), 
                      students = list(names = c("joe", "hexuan"), 
                                      ages = c(23, 25), 
                                      height = c(64, 64)))
```

Both of our lists are *named* lists (the best kind to work on). When we print those out, we can see those `$` printing out in the console. Let's see how they work:

```{r}
list_example$names

list_of_lists$professors$names
```

Let's chat about that `$`. R has a few special functions and this is one of them! Functions like `$`, `[`, and `[[` help us to access information inside of objects.

Here's how they work:

```{r}
list_example["names"]
list_example[1]

list_of_lists[["professors"]]
list_of_lists[[1]]
```

And just to reiterate that these really are nothing more than functions, check this out:

```{r}
`$`(list_example, names)

`[`(list_example, "names")

`[[`(list_of_lists, "professors")
```

And just to show you something silly:

```{r}
list_of_lists |> 
  getElement(1) |>
  getElement("names")
```

What do you think is the easiest thing to use?

## Data Frames

Now is the point in which R gets wacky. In the most simple way, a data frame is a special type of list; really, a data frame is a list with some strict rules. Chief among those rules is that every element of the data frame needs to contain the same number of items in each column; in other words, there needs to be the same number of rows.

```{r}
df_example <- data.frame(name = c("seth", "sharif",
                                  "joe", "hexuan", 
                                  "hermalena"), 
                         age = c(30, 30, 23, 25, 30), 
                         height = c(76, 76, 64, 64, 68), 
                         role = c("professor", "professor", 
                                  "student", "student", 
                                  "student_services"))

df_example

str(df_example)
```

Let's learn some helpful functions for looking at this data frame.

## Working With Data

Let's start with those square brackets and dollar signs.

Remember the dollar sign will extract out exactly 1 variable:

```{r}
df_example$age
```

If we want to *select* multiple variables, we can use the square brackets

```{r}
df_example[, c("age", "role")]
```

You could do the same thing with numbers:

```{r}
df_example[, c(2, 4)]
```

But...that is a dangerous game to play.

We can add more information into those brackets. They work like this:

```
df_example[row, column]
```

So if we only want to extract out the first row, we could do something like this:

```{r}
df_example[1, ]
```

Or, the first two rows:

```{r}
df_example[1:2, ]
```

We can also *filter* within the brackets.

```{r}
df_example[df_example$role == "student", ]
```

How would you filter for people over the age of 28?

You aren't limited to single filters! You can use the `&` sign and the `|` sign to specify and/or conditions.

```{r}
df_example[df_example$role == "student" & 
             df_example$name == "joe", ]
```

Let's do a little bit of grouping to our data now. We need to turn to some functions here to make this work. This function will be a little bit more complex than what we have seen, but we can work through it.

Let's just start with a simple mean of age:

```{r}
mean(x = df_example$age)
```

Now let's say that we want to group our data by the `role` variable and then get the average age for the two groups that remain. To do that, we can use the `aggregate` function.

```{r}
aggregate(x = df_example$age,
          by = list(group = df_example$role), 
          mean)

aggregate(age ~ role, 
          data = df_example, 
          mean)
```

Cool, but can we filter out everyone over 72" in height?

## Helpful Resources

The <a href="https://swirlstats.com/students.html">swirl</a> package is absolutely top notch!

The <a href="https://www.burns-stat.com/documents/books/the-r-inferno/">R Inferno</a> is old, but will never age.

<a href="https://www.statmethods.net/index.html">Quick-R</a> has been around forever. Unfortunate that DataCamp got involved with it.

The <a href="https://rseek.org/">rseek</a> page might be helpful.

You can waste time and get inspiration from <a href="https://www.r-bloggers.com/">R-bloggers</a>.

Finally, <a href="https://stackoverflow.com/">stackoverflow</a> has the answer to every question you might have.